Model for DES Type Ia Supernova Cosmology Analysis

### 2016-05-11 tex/toy.tex
a version for sam to look at.

\section{Likelihood}
\label{likelihood:sec}

The ambition is to construct a model to determine the likelihood of DES-like
supernova data.  We start with a toy model and plan to evolve its complexity to
have the fidelity necessary for the DES analysis.
The conceptual structure of the model is shown in Figure~\ref{pgm:fig}. 
\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=6.5in]{../results//nodes_pgm.pdf} 
   \caption{Probabilistic Graphical Model for the SN~Ia analysis.  
   \label{pgm:fig}}
\end{figure}

In the toy model there are three types of data that
may be observed for each supernova:
\begin{itemize}
\item $c_o$: Counts at peak brightness, one or more independent measurements.
\item $z_o$: Redshift.
\item $T_o$: Transient type.
\end{itemize}

@@

\subsection{Cosmology}
The Flat $w$CDM cosmological parameters are drawn from top-hat distributions
\begin{align}
\Omega_M & \sim  {\mathcal{U}}(0,1)\\
w_0 & \sim \mathcal{U}(-1.5, -0.1).
\end{align}

\subsection{Calibration}
Photometric zeropoints for each band.
\begin{equation}
Z \sim \mathcal{N}(0,\frac{\ln{10}}{2.5}0.01)
\end{equation}

\subsection{SNIa Rate}
Type~Ia SN rate. In the analysis only relative rates between different transients are important.
We take the rates relative to the SN~Ia rate
\begin{equation}
r_{SNIa} = 1.
\end{equation}

\subsection{SNII Rate}
Type~II SN rate, relative to the SN~Ia rate
\begin{equation}
r_{SNII} \sim \mathcal{U}(0.25, 4).
\end{equation}

\subsection{SNIa Luminosity}
The mean luminosity and intrinsic dispersion for SNe~Ia. In fact work with log-luminosity, intrinsic
dispersion in mag units.
\begin{align}
\ln{L}_{SNIa} & \sim \mathcal{N}(\ln{1}, 0.02) \\
\sigma_{SNIa} & \sim \ln{\mathcal{N}}(\ln{0.1},0.1).
\end{align}

\subsection{SNII Luminosity}
The mean luminosity and intrinsic dispersion for SNe~II. In fact work with log-luminosity, intrinsic
dispersion in mag units.
\begin{align}
\ln{L}_{SNII} & \sim \mathcal{N}(\ln{0.5}, 0.02) \\
\sigma_{SNII} & \sim \ln{\mathcal{N}}(\ln{0.4},0.1).
\end{align}

>>>>

\section{Likelihood and Posterior}
\label{likelihood:sec}

The ambition is to construct a model to determine the likelihood of DES-like
supernova data. 
The conceptual structure of the model is shown in Figure~\ref{pgm:fig}. 
\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=6.5in]{../results//hdpgm.pdf} 
   \caption{Probabilistic Graphical Model for the SN~Ia analysis.  
   \label{pgm:fig}}
\end{figure}

The model has three types of observables $O$ that
may be observed for each supernova:
\begin{itemize}
\item $c_o$: Counts of the photometric measurements (light curves).
\item $s_o$: Counts of the spectroscopic measurements.
\item $z_o$: Redshift.
\item $T_o$: Transient type.
\end{itemize}
Measurements of an observable are specified with a subscript ``o''.

There are top-level  sets of parameters $M$ that correspond to photometric calibration, cosmology,
host-galaxy properties, and transient properties.
There are two sets of intermediary parameters $M_I$ that correspond to luminosity $L$ and type $\tau$.

The likelihood for observations $O_o$ is expressed as
\begin{align*}
\mathcal{L}(O_o, sample| M, M_I) & = \frac{\epsilon(O_o, M, M_I) p(O_o|M, M_I)}{\int  \epsilon(O, M, M_I) p(O|M, M_I)dO},
\end{align*}
where $p(O|M, M_I)$ is the pdf of all possible observables, and $\epsilon(O, M, M_I) $
is the probability those observables will be included in the experimental analysis.

The dependencies of the parameters are charted through the PGM to give for transients with a type measurement
\begin{align*}
p(O| M, M_I) & = P(T | \tau) p(z | M) p(c | L, M)
\end{align*}
and for transients without a type measurement (not relevant for the SN~Ia-only analysis but included for completeness)
\begin{align*}
p(O| M, M_I) & = p(z | M) p(c | L, M),
\end{align*}
where the observable measurements are taken to be independent.

The efficiency $\epsilon(O, M, M_I) $ depends on the properties of the experiment and the sample selection criteria:
the transient is discovered, satisfies sample selection, is selected for spectroscopic typing, the typing
is successful, and that the type is SN~Ia:
\begin{align*}
\epsilon(O, M, M_I)  & = P(SNIa, typed, spec, sample, discovered |O, M, L, \tau)\\
& = P(SNIa|  typed, \tau)P(typed | spec, s, M, M_I)P(spec | sample, c) \\
& \times P(sample| discovered, c)P(discovered |c),
\end{align*}
where transient discovery, sample selection, and spectroscopic follow-up are based on observed photometry.

The posterior is given by Bayes theorem
\begin{align*}
p(M, M_i | O_o) & \propto \mathcal{L}(O_o| M, M_I) p(M_I|M) p(M)\\
 & \propto \mathcal{L}(O_o| M, M_I) p(L| \tau, M) p(\tau|M) p(M).
\end{align*}

\section{Probabilities}
Section~\ref{likelihood:sec} introduced a series of probabilities based on the structure of our model, which are now described
in further detail.

\subsection{$p(M)$: Priors on the Top-Level Parameters}
\begin{itemize}
\item
The Flat $w$CDM cosmological parameters are drawn from top-hat distributions
\begin{align}
\Omega_M & \sim  {\mathcal{U}}(0,1)\\
w_0 & \sim \mathcal{U}(-1.5, -0.1).
\end{align}

\item
Photometric zeropoints for each band.
\begin{equation}
Z \sim \mathcal{N}\left(0,\frac{\ln{10}}{2.5}0.01\right)
\end{equation}

\item
There are a number of parameters that describe the supernova populations.

The luminosity of SNe~Ia as a function of time and wavelength are described by the SALT2 model and
its relation to absolute magnitude.
The mean luminosity and intrinsic magnitude dispersion for SNe~Ia are
\begin{align}
\ln{L}_{SNIa} & \sim \mathcal{N}(\ln{1}, 0.02) \\
\sigma_{SNIa} & \sim \ln{\mathcal{N}}(\ln{0.1},0.1).
\end{align}
and for the moment we don't worry about priors on the distribution of $\alpha$ and $\beta$.

For the cosmology analysis only the relative SN~Ia rate between different transients are important.
We take the rates relative to the SN~Ia rate
\begin{equation}
r_{SNIa} = 1.
\end{equation}
This is sufficient for the SN~Ia sample analysis.

When considering the photometric sample the model needs to consider non-Ia populations.
The mean luminosity and intrinsic dispersion for non-Ia in a simplified toy model
\begin{align}
\ln{L}_{SNII} & \sim \mathcal{N}(\ln{0.5}, 0.02) \\
\sigma_{SNII} & \sim \ln{\mathcal{N}}(\ln{0.4},0.1),
\end{align}
and make the non-Ia rate a stochastic parameter
\begin{equation}
r_{SNII} \sim \mathcal{U}(0.25, 4).
\end{equation}

\end{itemize}

\subsection{$p(M)$: Priors on the Top-Level Parameters}

@@
